Spike-and-Slab Generalized Additive Models and Scalable Algorithms for High-Dimensional Data

### 2021-09-11 Manuscript/SS_GAM.tex
Finish Revision

\title{Spike-and-Slab Generalized Additive Models and Fast Algorithms
for High-Dimensional Data}

@@

\usetikzlibrary{shapes.geometric, arrows, positioning, calc, matrix, backgrounds, fit}
\newcommand{\bs}[1]{\boldsymbol{#1}}
\newcommand{\tp}{\text{pen}}

@@

Many modern biomedical research, e.g.~sequencing data analysis, electric
health record data analysis, require special treatment of
high-dimensionality, commonly known as \(p >> n\) problem. There is
extensive literature on high-dimensional linear models via penalized
models or Bayesian hierarchical models, see Mallick and Yi
\citep{Mallick2013} for review. These models are built upon a
restrictive and unrealistic assumption, linearity. In classical
statistical modelling, many strategies and models are proposed to relax
the linearity assumption with various degrees of complexity. For
example, variable categorization is a simple and common practice in
epidemiology, but suffers from power and interpretation issues. More
complex models to address nonlinear effects include random forest and
other so-called ``black box'' models \citep{Breiman2001}. These models
are not particularly attractive in biomedical studies, as they lack the
transparency to understand the etiology of diseases. In addition, how to
generalized these models to the high-dimensional setting remains
unclear.

@@

There exists some proposals to generalize the classical GAM to
accommodate high-dimensional applications. The regularized models,
branching out from regularized linear models, are used to fit GAMs by
accounting for the group structure introduced when expanding smoothing
functions. Ravikumar et al. \citep{Ravikumar2009} extended the grouped
lasso \citep{Yuan2006} to additive models; Huang et al.
\citep{Huang2010} further developed adaptive grouped lasso for additive
models; Wang et al. \citep{Wang2007} and Xue \citep{Xue2009}
respectively applied grouped SCAD penalty \citep{Fan2001} to the
additive models. Recent Bayesian hierarchical models are also used in
the context of high-dimensional additive models. Various group
spike-and-slab priors combining with computationally intensive Markov
chain Monte Carlo (MCMC) algorithms \citep{Xu2015, Yang2020} are
proposed, where the application on AMs are by-products. Bai et al.
\citep{Bai2020} was the first to apply group spike-and-slab lasso prior
to Gaussian additive models(AMs) using a fast optimization algorithm,
and further generalized the framework to GAMs\citep{Bai2021}. As these
methods focus on addressing the sparsity, it is very likely to overly
penalize the basis function coefficients and produce inaccurate
predictions and curve interpolation, particularly when complex signals
are assumed and large number of knots are used. \citep{Scheipl2013} In
addition, the proposed methods adapt an `all-in-all-out' strategy,
i.e.~either including or excluding the variable completely, rendering no
space for bi-level selection. Scheipl et al. \citep{Scheipl2012}
proposed an ad hoc spike-and-slab structure prior,
normal-mixture-of-inverse gamma prior, that address the previous
challenges. But the model fitting relies on computational intensive MCMC
algorithmic and creates scalability concern. It would be of special
interest to develop a fast, flexible and accurate generalized additive
model.

To address these challenges, we propose a novel Bayesian hierarchical
generalized additive model (BHAM) for high dimensional data analysis.
Specifically, we incorporate smoothness penalties in the model via
re-parameterization of the smoothing function to avoid overly shrinking
basis function coefficients. Smoothness penalties are commonly
implemented in GAMs through smoothing regression splines. They are
quadratic norms of the coefficients and allow locally adaptive penalties
on each smoothing function. The smoothness penalty conditioning on a
smoothness penalty parameter \(\lambda_j\) is a function of the
integration of the second derivative of the spline function, expressed
mathematically as \begin{equation}\label{eq:smoothpen}
  \text{pen}\left[f_j(x)\right] = \lambda_j \int f^{\prime\prime}_j(x)dx = \lambda_j \boldsymbol{\beta}_j^T \boldsymbol{S}_j \boldsymbol{\beta}_j ,
\end{equation} where \(\boldsymbol{S}_j\) is a known smoothness penalty
matrix, and \(\boldsymbol{\beta}_j\) is the basis function coefficients.
Smoothness penalties are also previously used in the spike-and-slab GAM
\citep{Scheipl2012} and the sparsity-smoothness penalty
\citep{Meier2009}. Moreover, incorporating the smoothness penalty allows
the separation of the linear space of smoothing functions, also know as
the null space, from the nonlinear space. We then impose a new bi-part
spike-and-slab spline prior on the smoothing functions for bi-level
selection such that the linear and nonlinear spaces of smoothing
functions can be selected separately. The prior setup encourages a
flexible solution, rending one of three possibilities: no effect, only
linear effect, or nonlinear effect of a predictor. In addition, two fast
computing optimization-based algorithms are developed for high and
ultrahigh-dimensional\citep{Fan2009} settings respectively.

The proposed framework, BHAM differs from previous spike-and-slab based
GAMs, i.e.~the spike-and-slab GAM \citep{Scheipl2012} and the SB-GAM
\citep{Bai2021} in many ways. First of all, the proposed spike-and-slab
spline prior is a spike-and-slab lasso type prior using independent
mixture double exponential distribution, compared to spike-and-slab GAM.
Spike-and-slab lasso priors provide computational convenience during
model fitting by using optimization algorithms instead of computational
intensive sampling algorithms. They make fitting high- and
ultrahigh-dimensional models more feasible without sacrificing
performance in prediction and variable selection. Secondly, the
optimization algorithms introduced in this article include EM-Coordinate
Descent algorithm and EM-Iterative weighted least square algorithm for
different utilities. The applications of the two aforementioned
algorithm are fruitful in Bayesian hierarchical models in
high-dimensional data analysis.
\citep{Yi2012, Rockova2014a, Rockova2018} Even though SB-GAM with a
group spike-and-slab lasso prior also uses EM-CD for model fitting, the
proposed EM-IWLS provides uncertainty measures that could be further
used for confidence bound calculation and hypothesis testing. EM-IWLS
complements the lack of inference potential from EM-CD in a median and
high dimensional setting. Furthermore, the proposed model addresses the
incapability of bi-level selection in SB-GAM. Lastly, the R
implementation of the Bayesian hierarchical additive model is available
publicly via BHAM at \url{https://github.com/boyiguo1/BHAM}, make
transnational science more accessible.

In Section \ref{sec:BHAM}, we establish the Bayesian hierarchical
generalized additive model, introduce the proposed spike-and-slab spline
priors, and describe the two fast-fitting algorithms. In Section
\ref{sec:sim}, we compare the proposed framework to state-of-art models,
mgcv, COSSO and sparse Bayesian GAM via monte carlo simulation studies.
Analyses of two metabolomics datasets using the proposed model are
presented in Section \ref{sec:real_data}. Conclusion and discussions are
given in Section \ref{sec:concl}.

@@

Following the GLM notation introduced in Section \ref{sec:intro}, we
have a generalized additive model and its matrix form,
\begin{equation}\label{eq:gam}
E(Y|\boldsymbol{x}) = g^{-1}(a + \sum\limits^p_{j=1}B_j(x_j)) = g^{-1}(a + \sum\limits^p_{j=1} \boldsymbol{\beta}_j^T \boldsymbol{X}_j),
\end{equation} with smoothing functions \(B_j(x_j)\) of the variable
\(x_j, j = 1, \dots, p.\) The basis function matrix, i.e.~the design
matrix derived from the smoothing function \(B_j(x_j)\), is denoted
\(\boldsymbol{X}_j\) for the variable \(x_j\). The dimension of the
design matrix depends on the choice of the smoothing function. For
example, a cubic spline function with \(k\) knots produces a \(k+2\)
elements column vector. Its corresponding smoothness penalty matrix is
denoted as \(\boldsymbol{S}_j\). \(\boldsymbol{\beta}_j\) denotes the
basis function coefficients for the \(j\)th variable, and the matrix
transposing operation is denoted with a superscript \(^T\). To note, the
proposed model can include parametric forms of variables in the model,
treating the parametric function as a special case of the smoothing
function, e.g.~\(B_j(x_j) = x_j\) with the smoothness penalty matrix
defined as \(\boldsymbol{S}_j = \begin{bmatrix}0\end{bmatrix}\).

To encourage proper smoothness of the smoothing functions, we adopt the
idea of smoothing penalties from smoothing spline models. The basic idea
is to set up a smoothing penalty, depending on the smoothness penalty
matrix via Equation \ref{eq:smoothpen}, in the posterior density
function, or least squared and likelihood function in the penalized
framework. However, the direct integration of smoothness penalties into
sparsity penalties is not obvious. Given the smoothness penalty matrix
is most often symmetric and positive semi-definite \footnote{ we limit
  the discussion to univariate smoothing function. However, the penalty
  matrix could possibly contain multiple matrices, e.g.~tensor product
  smoother.}, it can be integrated into the design matrix
\(\boldsymbol{X}\) via a re-parameterization. The re-parameterization
follows Marra and Wood \citep{Marra2011}, in which eigen-decomposed the
penalty matrix
\(\boldsymbol{S} = \boldsymbol{U} \boldsymbol{D} \boldsymbol{U}^T\),
where the matrix \(\boldsymbol{D}\) is diagonal with the eigenvalues
arranged in the ascending order. To note, \(\boldsymbol{D}\) can contain
\(0\) elements on the diagonal, associated with the null space of the
smooth function. We further write the orthonormal matrix
\(\boldsymbol{U}\), containing the eigenvectors as column, as an (row)
augmentation of two sub-matrices
\(\boldsymbol{U} \equiv \begin{bmatrix}\boldsymbol{U}^0 : \boldsymbol{U}^{\text{pen}}\end{bmatrix}\),
where \(\boldsymbol{U}^{\text{pen}}\) contains the eigenvectors (as
columns) for the non-zero eigenvalues, i.e.~the penalized space, and
\(\boldsymbol{U}^0\) contains the eigenvectors for the null space. We
multiply the basis function matrix \(\boldsymbol{X}\) with the
orthonormal matrix \(\boldsymbol{U}\) for the new design matrix
\(\boldsymbol{X}^\ast = \boldsymbol{X} \boldsymbol{U} \equiv \begin{bmatrix} \boldsymbol{X}^0 : \boldsymbol{X}^{\text{pen}} \end{bmatrix}\).
An additional scaling step is imposed on \(\boldsymbol{X}^{\text{pen}}\)
by the non-zero eigenvalues of \(\boldsymbol{D}\) such that the new
basis function matrix \(\boldsymbol{X}^\ast\) is scale-invariant
\footnote{ I know this work for tensor spline (see \citep{Wood2006}),
  but I am not sure if it can be generalized to all smoothing functions}.
For notation convenience, we denote
\(\boldsymbol{X}_j \equiv \begin{bmatrix} \boldsymbol{X}_j^0 : \boldsymbol{X}_j^{\text{pen}} \end{bmatrix}\)
as the after re-parameterized basis function matrix for the \(j\)th
variable hereafter. Similarly, a spline function can be expressed in the
matrix form \[
f_j(x_j) = f_j^0(x_j) + f_j^\text{pen}(x_j) = \boldsymbol{\beta_j^0}^T \boldsymbol{X}_j^0 + \boldsymbol{\beta_j^\text{pen}}^T \boldsymbol{X}_j^\text{pen},
\] and the generalized additive model in Equation (\ref{eq:gam}) be
\begin{equation}\label{eq:gam-repa}
E(Y|\boldsymbol{x}) =  g^{-1}(a + \sum\limits^p_{j=1} \boldsymbol{\beta}_j^T \boldsymbol{X}_j) = g^{-1}\left[a + \sum\limits^p_{j=1} ({\beta_j^0}^T \boldsymbol{X}_j^0 + {\beta_j^\text{pen}}^T \boldsymbol{X}_j^\text{pen})\right].
\end{equation}

The re-parameterization sets up the foundation of the proposed model,
which has three-fold benefits. First of all, as previously mentioned,
the re-parameterization integrated the smoothness penalty into the
design matrix, and hence the penalty can be expressed in the posterior
density function implicitly in addition to the sparse penalty for
variable and function selection. Secondly, the eigendecomposition of the
smooth penalty isolates the linear space of a smooth function from the
nonlinear space, improving the feasibility of bi-level variable/function
selection and estimation inference. The eigendecomposition also
introduced the linearly independent column spaces of the smooth function
matrix, which allows imposing independent priors on the coefficients.
This reduces the computational complexity that multivariate priors
induces, and greatly broadens the choices of priors and further model
choices. Last but not least, the scaling step of
\(\boldsymbol{X}^\text{pen}\) further simplifies the choice of penalty
priors, from column-specific penalties to a unified penalty. This step
is also necessary for the proposed spike-and-slab spline prior.
Conventionally, the spike-and-slab priors require scaling of predictors
\footnote{TODO: insert a citation.}

@@

The family of spike-and-slab (SS) priors regression models dominates
Bayesian high-dimensional analysis for its utility in outcome prediction
and variable selection. We defer to Bai et al. \citep{Bai2021Review} for
an in-depth introduction to spike-and-slab priors. To summary,
spike-and-slab priors are a family of mixture distributions that
comprises a skinny spike density \(f_{spike}(\cdot)\) for weak signals
and a flat slab density \(f_{slab}(\cdot)\) for large signals,
mathematically \[
 \beta|\gamma \sim (1-\gamma)f_{spike}(\beta) + \gamma f_{slab}(\beta).
\] The most distinct feature of the SS priors is that it is conditioned
on a latent binary variable \(\gamma\) that indicates whether the
variable \(x\) is included in the model. There are various
spike-and-slab priors depending on the choice for the spike density
\(f_{spike}(\cdot)\)\footnote{TODO: should I change to notation of \(f\)
  to not confound with the smooth functions?} and the slab density
\(f_{slab}(\cdot)\), see George and McCulloch
\citep{George1993, George1997}; Chipman \citep{Chipman1996} for grouped
variables; Brown et al. \citep{Brown1998} for multivariate outcomes;
Ishwaran and Rao \citep{Ishwaran2005}; Clyde and George
\citep{Clyde2004} and reference therein.

The major criticism of early spike-and-slab priors receives is being
computationally prohibitive. \citep{Bai2021Review} Since then, many
studies focus on alleviating the computational burden that sampling
algorithms bear, which include EMVS \citep{Rockova2014a} and
Spike-and-slab Lasso \citep{Rockova2018b, Rockova2018}. Particularly,
the development of Spike-and-slab Lasso substantially improves the
scalability of SS models, setting up the theoretical foundation for
generalized models in -omics data analysis
\citep{Tang2017a, Tang2017, Tang2018, Tang2019}. The spike-and-slab
lasso prior is composed of two double exponential distributions with
mean 0 and different dispersion parameters, \(0 < s_0 < s_1\),
mathematically, \begin{equation} \label{eq:ssl}
\beta | \gamma \sim (1-\gamma)DE(0, s_0) + \gamma DE(0, s_1), 0 < s_0 < s_1.
\end{equation} The SSL also mitigated the problem of EMVS where the weak
signals are not shrink to zero, favored in high-dimensional data
analysis. We noticed that Bai \citep{Bai2021} introduces a
spike-and-slab group prior in the GAM framework, where the densities of
the spike and slab components take the group lasso density
\citep{Xu2015}. Bai \citep{Bai2021} takes an ``all-in-all-out'' strategy
for function selection.

@@

We introduce a novel prior for GAMs, particularly for high-dimensional
nonlinear effect modeling with bi-level selection. The proposed prior
extends from the spike-and-slab lasso prior described in Equation
\ref{eq:ssl}. Given the reparameterized deign matrix
\(\boldsymbol{X}_j = \begin{bmatrix} \boldsymbol{X}_j^0 : \boldsymbol{X}_j^\text{pen}\end{bmatrix}^T\)
for the \(j\)th variable, we impose a two-part SSL prior to the
coefficients \(\boldsymbol{\beta}_j\). Specifically, we impose
independent group priors on the null space coefficients and on the
penalized space coefficients respectively,

\begin{align}\label{eq:bham_ssl}
  \beta^0_{j} |\gamma^0_{j},s_0,s_1 &\sim DE(0,(1-\gamma^0_{j}) s_0 + \gamma^0_{j} s_1)\\
  \beta^\text{pen}_{jk} | \gamma^\text{pen}_{j},s_0,s_1 &\sim DE(0,(1-\gamma^\text{pen}_{j}) s_0 + \gamma^\text{pen}_{j} s_1), 
\end{align}

where \(\gamma^0_{j}\in\{0,1\}\) and
\(\gamma^\text{pen}_{j}\in \{0,1\}^{K_j}\) are latent indicator
variables, indicating if the model includes the linear effect and the
nonlinear effect of the \(j\)th variable respectively. ; \(s_0\) and
\(s_1\) are scale parameters, assuming \(0 < s_0 < s_1\) and given.
These scale parameters \(s_0\) and \(s_1\) can be treated as tuning
parameters and optimized via cross-validation. A discussion of how to
choose the scale parameters comes in Section \ref{sec:tune}. To note,
this prior differs from previous group spike-and-slab lasso priors
\citep{Tang2018, Tang2019}, where the \(\boldsymbol{\beta}^0_j\) and
\(\boldsymbol{\beta}^\text{pen}_j\) shares the same indicator variables
\(\gamma_j^0\), \(\gamma_j^\text{pen}\) respectively. It is possible to
relax the assumption such that each coefficient \(\beta\) to have its
own latent indicator \(\gamma\), but at the cost ofcomplicating the
bi-level function selection. Conversely, it is possible to add a more
restrictive assumption on the priors, assuming that one indicator
variable decides the inclusion of both the linear effect and nonlinear
effect, i.e.~\(\gamma_j = \gamma^0_j = \gamma^\text{pen}_j\). This
converges\footnote{Is ``becomes'' better?} to the SB-GAM
\citep{Bai2021}.

The reparameterization introduced in Section \ref{sec:BHAM} grants the
validity of the proposed prior. First of all, the smooth function bases
are linear dependent to each other based on the definitions, and ask for
extra attention. The eigeondecomposition remedies the problem. Hence our
prior can be set to be conditionally independent. Secondly, the
eigenvalue scaling provides a panacea to allow unified scale parameters
for all bases of all smoothing functions of variables, where scaling is
mandate.

>>>>

\title{Spike-and-Slab Generalized Additive Models and Scalable
Algorithms for High-Dimensional Data}

@@

\pgfplotsset{compat=1.18}
\usetikzlibrary{shapes.geometric, arrows, positioning, calc, matrix, backgrounds, fit}
\newcommand{\bs}[1]{\boldsymbol{#1}}
\newcommand{\tp}{\text{pen}}
\newcommand{\pr}{\text{Pr}}

@@

Many modern biomedical research, e.g.~sequencing data analysis, electric
health record data analysis, require special treatment of
high-dimensionality, commonly known as \(p >> n\) problem. There is
extensive literature on high-dimensional linear models via penalized
models or Bayesian hierarchical models, see Mallick and Yi
\citep{Mallick2013} for review. These models are built upon a
restrictive and unrealistic assumption, linearity. In classical
statistical modeling, many strategies and models are proposed to relax
the linearity assumption with various degrees of complexity. For
example, variable categorization is a simple and common practice in
epidemiology, but suffers from power and interpretation issues. More
complex models to address nonlinear effects include random forest and
other so-called ``black box'' models \citep{Breiman2001}. These models
are not particularly attractive in biomedical studies, as they lack the
transparency to understand the etiology of diseases. In addition, how to
generalized these models to the high-dimensional setting remains
unclear.

@@

There exists some proposals to generalize the classical GAM to
accommodate high-dimensional applications. The regularized models,
branching out from group regularized linear models, are used to fit GAMs
by accounting for the structure introduced when expanding smoothing
functions. Ravikumar et al. \citep{Ravikumar2009} extended the grouped
lasso \citep{Yuan2006} to additive models (AMs); Huang et al.
\citep{Huang2010} further developed adaptive grouped lasso for additive
models; Wang et al. \citep{Wang2007} and Xue \citep{Xue2009}
respectively applied grouped SCAD penalty \citep{Fan2001} to additive
models. Recent Bayesian hierarchical models are also used in the context
of high-dimensional additive models. Various group spike-and-slab priors
combining with computationally intensive Markov chain Monte Carlo (MCMC)
algorithms \citep{Xu2015, Yang2020} are proposed, where the application
on AMs are by-products. Bai et al. \citep{Bai2020} was the first to
apply group spike-and-slab lasso prior to Gaussian AMs using a fast
optimization algorithm, and further generalized the framework to
GAMs\citep{Bai2021}. As these methods focus on addressing the sparsity,
it is very likely to overly penalize the basis function coefficients and
produce inaccurate predictions and curve interpolation, particularly
when complex signals are assumed and large number of knots are used.
\citep{Scheipl2013} In addition, these methods adapt an `all-in-all-out'
strategy, i.e.~either including or excluding the variable completely,
rendering no space for bi-level selection. Scheipl et al.
\citep{Scheipl2012} proposed an ad hoc spike-and-slab structure prior,
normal-mixture-of-inverse gamma prior, that address the previous
challenges. But the model fitting relies on computational intensive MCMC
algorithmic and creates scalability concern. It would be of special
interest to develop a fast, flexible and accurate generalized additive
model framework.

To address these challenges, we propose a novel Bayesian hierarchical
generalized additive model (BHAM) for high dimensional data analysis.
Specifically, we incorporate smoothing penalties in the model via
re-parameterization of the smoothing function to avoid overly shrinking
basis function coefficients. Smoothing penalties are commonly
implemented in the classical GAMs through smoothing regression splines.
They are quadratic norms of the coefficients and allow locally adaptive
penalties on each smoothing function. A smoothing penalty conditioning
on a smoothing parameter \(\lambda_j\) is a function of the integration
of the second derivative of the spline function, expressed
mathematically as \begin{equation}\label{eq:smoothpen}
  \text{pen}\left[f_j(x)\right] = \lambda_j \int f^{\prime\prime}_j(x)dx = \lambda_j \boldsymbol{\beta}_j^T \boldsymbol{S}_j \boldsymbol{\beta}_j ,
\end{equation} where \(\boldsymbol{S}_j\) is a known smoothing penalty
matrix, and \(\boldsymbol{\beta}_j\) are the basis function
coefficients. Smoothing penalties re also previously used in the
spike-and-slab GAM \citep{Scheipl2012} and the sparsity-smoothness
penalty \citep{Meier2009}. Moreover, incorporating the smoothing penalty
allows the separation of the linear space of smoothing functions, also
know as the null space, from the nonlinear space. We then impose a new
bi-part spike-and-slab spline prior on the smoothing functions for
bi-level selection such that the linear and nonlinear spaces of
smoothing functions can be selected separately. The prior setup
encourages a flexible solution, rending one of three possibilities: no
effect, only linear effect, or nonlinear effect of a predictor. In
addition, two fast computing optimization-based algorithms are developed
for high and ultrahigh-dimensional\citep{Fan2009} settings respectively.

The proposed framework, BHAM, differs from previous spike-and-slab based
GAMs, i.e.~the spike-and-slab GAM \citep{Scheipl2012} and the SB-GAM
\citep{Bai2021} in many ways. First of all, the proposed spike-and-slab
spline prior is a spike-and-slab lasso type prior using independent
mixture double exponential distribution, compared to spike-and-slab GAM.
Spike-and-slab lasso priors provide computational convenience during
model fitting by using optimization algorithms instead of computational
intensive sampling algorithms. They make fitting high- and
ultrahigh-dimensional models more feasible without sacrificing
performance in prediction and variable selection. Secondly, the
optimization algorithms introduced in this article include EM-Coordinate
Descent algorithm and EM-Iterative weighted least square algorithm for
different utilities. The applications of the two aforementioned
algorithm are fruitful in Bayesian hierarchical models in
high-dimensional data analysis.
\citep{Yi2012, Rockova2014a, Rockova2018} Even though SB-GAM with a
group spike-and-slab lasso prior also uses EM-CD for model fitting, the
proposed EM-IWLS provides uncertainty measures that could be further
used for confidence bound calculation and hypothesis testing. EM-IWLS
complements the lack of inference potential from EM-CD in a median and
high dimensional setting. Furthermore, the proposed model addresses the
incapability of bi-level selection in SB-GAM. Lastly, the R
implementation of the Bayesian hierarchical additive model is available
publicly via BHAM at \url{https://github.com/boyiguo1/BHAM}, make
transnational science more accessible.

In Section \ref{sec:BHAM}, we establish the Bayesian hierarchical
generalized additive model, introduce the proposed spike-and-slab spline
priors, and describe the two fast-fitting algorithms. In Section
\ref{sec:sim}, we compare the proposed framework to state-of-art models,
mgcv, COSSO and sparse Bayesian GAM via monte carlo simulation studies.
Analyses of two metabolomics datasets are presented in Section
\ref{sec:real_data}. Conclusion and discussions are given in Section
\ref{sec:concl}.

@@

Following the GLM notation introduced in Section \ref{sec:intro}, we
have a generalized additive model and its matrix form,
\begin{equation}\label{eq:gam}
E(Y|\boldsymbol{x}) = g^{-1}(a + \sum\limits^p_{j=1}B_j(x_j)) = g^{-1}(a + \sum\limits^p_{j=1} \boldsymbol{\beta}_j^T \boldsymbol{X}_j),
\end{equation} with smoothing functions \(B_j(x_j)\) of the variable
\(x_j, j = 1, \dots, p.\) The basis function matrix, i.e.~the design
matrix derived from the smoothing function \(B_j(x_j)\), is denoted
\(\boldsymbol{X}_j\) for the variable \(x_j\). The dimension of the
design matrix depends on the choice of the smoothing function. For
example, a cubic spline function with \(k\) knots produces a \(k+3\)
elements column vector. Its corresponding smoothing penalty matrix is
denoted as \(\boldsymbol{S}_j\). \(\boldsymbol{\beta}_j\) denotes the
basis function coefficients for the \(j\)th variable, and the matrix
transposing operation is denoted with a superscript \(^T\). To note, the
proposed model can include parametric forms of variables in the model,
treating the parametric function as a special case of the smoothing
function, e.g.~\(B_j(x_j) = x_j\) with the smoothing penalty matrix
defined as \(\boldsymbol{S}_j = \begin{bmatrix}0\end{bmatrix}\).

To encourage proper smoothing of the functions, we adopt the idea of
smoothing penalties from smoothing spline models. The basic idea is to
set up a smoothing penalty, described in Equation (\ref{eq:smoothpen}),
in the prior density function, or least squared and likelihood function
in the penalized framework. However, the direct integration of smoothing
penalties into sparsity penalties is not obvious. Marra and Wood
\citep{Marra2011} proposed a re-parameterization procedure to
accommodate the smoothing penalty implicitly. Given the smoothing
penalty matrix is most often symmetric and positive semi-definite for
univariate smoothing functions, we apply eigendecomposition on the
penalty matrix
\(\boldsymbol{S} = \boldsymbol{U} \boldsymbol{D} \boldsymbol{U}^T\) ,
where the matrix \(\boldsymbol{D}\) is diagonal with the eigenvalues
arranged in the ascending order. To note, \(\boldsymbol{D}\) can contain
\(0\) elements on the diagonal, associated with the null space of the
smoothing function. We further write the orthonormal matrix
\(\boldsymbol{U} \equiv \begin{bmatrix}\boldsymbol{U}^0 : \boldsymbol{U}^{\text{pen}}\end{bmatrix}\)
containing the eigenvectors as columns in cthe orresponding order. That
is \(\boldsymbol{U}^0\) contains the eigenvectors with zero eigenvalues
for the null space and \(\boldsymbol{U}^{\text{pen}}\) contains the
eigenvectors (as columns) for the non-zero eigenvalues, i.e.~the
penalized space. We multiply the basis function matrix
\(\boldsymbol{X}\) with the orthonormal matrix \(\boldsymbol{U}\) for
the new design matrix
\(\boldsymbol{X}^\ast = \boldsymbol{X} \boldsymbol{U} \equiv \begin{bmatrix} \boldsymbol{X}^0 : \boldsymbol{X}^{\text{pen}} \end{bmatrix}\).
An additional scaling step is imposed on \(\boldsymbol{X}^{\text{pen}}\)
by the non-zero eigenvalues of \(\boldsymbol{D}\) such that the new
basis function matrix \(\boldsymbol{X}^\ast\) can receive uniform
penalty on each of its dimensions. With slight abuse of the notation, we
denote
\(\boldsymbol{X}_j \equiv \begin{bmatrix} \boldsymbol{X}_j^0 : \boldsymbol{X}_j^{\text{pen}} \end{bmatrix}\)
as the basis function matrix for the \(j\)th variable after the
re-parameterization. A spline function can be expressed in the matrix
form \[
B_j(x_j) = B_j^0(x_j) + B_j^\text{pen}(x_j) = \boldsymbol{\beta_j^0}^T \boldsymbol{X}_j^0 + \boldsymbol{\beta_j^\text{pen}}^T \boldsymbol{X}_j^\text{pen},
\] and the generalized additive model in Equation (\ref{eq:gam}) be
\begin{equation}\label{eq:gam-repa}
E(Y|\boldsymbol{x}) =  g^{-1}(a + \sum\limits^p_{j=1} \boldsymbol{\beta}_j^T \boldsymbol{X}_j) = g^{-1}\left[a + \sum\limits^p_{j=1} (\boldsymbol{\beta_j^0}^T \boldsymbol{X}_j^0 + {\boldsymbol{\beta}_j^\text{pen}}^T \boldsymbol{X}_j^\text{pen})\right].
\end{equation}

The re-parameterization step sets up the foundation of the proposed
model and provides three-fold benefits. First of all, the
re-parameterization integrates the smoothing penalty into the design
matrix, and encourages models to properly smooth the nonlinear function
in addition to the sparse penalty for function selection. Secondly, the
eigendecomposition of the smoothing penalty allows the isolation of the
linear from the nonlinear space, improving the feasibility of bi-level
function selection and inference. The eigendecomposition facilitates the
construction of orthonormal design matrix, which makes imposing
independent priors on the coefficients possible. This reduces the
computational complexity compared to using a multivariate priors, and
greatly broadens the choices of priors and further model choices. Last
but not least, the scaling step of \(\boldsymbol{X}^\text{pen}\) further
simplifies the choice of priors, from column-specific priors to a
unified prior.

@@

The family of spike-and-slab (SS) priors regression models dominates
Bayesian high-dimensional analysis for its utility in outcome prediction
and variable selection. We defer to Bai et al. \citep{Bai2021Review} for
an in-depth introduction to spike-and-slab priors. To summarize,
spike-and-slab priors are a family of mixture distributions that
comprises a skinny spike density \(f_{\text{spike}}(\cdot)\) for weak
signals and a flat slab density \(f_{\text{slab}}(\cdot)\) for strong
signals, mathematically \[
 \beta|\gamma \sim (1-\gamma)f_{\text{spike}}(\beta) + \gamma f_{\text{slab}}(\beta).
\] The most distinct feature of SS priors is that it is conditioned on a
latent binary variable \(\gamma\) that indicates whether the variable
\(x\) is included in the model. There are various spike-and-slab priors
depending on the choice for the spike density
\(f_{\text{spike}}(\cdot)\) and the slab density
\(f_{\text{slab}}(\cdot)\), see George and McCulloch
\citep{George1993, George1997}; Chipman \citep{Chipman1996} for grouped
variables; Brown et al. \citep{Brown1998} for multivariate outcomes;
Ishwaran and Rao \citep{Ishwaran2005}; Clyde and George
\citep{Clyde2004} and reference therein.

The major criticism of early spike-and-slab priors receive is being
computationally prohibitive. \citep{Bai2021Review} Since then, many
studies focus on alleviating the computational burden that sampling
algorithms bear, which include EMVS \citep{Rockova2014a} and
Spike-and-slab Lasso \citep{Rockova2018b, Rockova2018}. Particularly,
the development of Spike-and-slab Lasso substantially improves the
scalability of SS models, setting up the theoretical foundation for
generalized models in -omics data analysis
\citep{Tang2017a, Tang2017, Tang2018, Tang2019}. The spike-and-slab
lasso prior is composed of two double exponential distributions with
mean 0 and different dispersion parameters, \(0 < s_0 < s_1\),
mathematically, \begin{equation} \label{eq:ssl}
\beta | \gamma \sim (1-\gamma)DE(0, s_0) + \gamma DE(0, s_1), 0 < s_0 < s_1.
\end{equation} The SSL also mitigates the problem of EMVS where the weak
signals are not shrink to zero, and hence is prefered in
high-dimensional data analysis. We notice that Bai \citep{Bai2021} is
the first to apply spike-and-slab lasso prior in the GAM framework,
where the densities of the spike and slab components take the group
lasso density \citep{Xu2015}. Bai \citep{Bai2021} takes an
``all-in-all-out'' strategy for function selection.

@@

We introduce a novel prior for GAMs, particularly for high-dimensional
nonlinear modeling with bi-level selection. The proposed prior extends
from the spike-and-slab lasso prior described in Equation
(\ref{eq:ssl}). Given the re-parameterized deign matrix
\(\boldsymbol{X}_j = \begin{bmatrix} \boldsymbol{X}_j^0 : \boldsymbol{X}_j^\text{pen}\end{bmatrix}\)
for the \(j\)th variable, we impose a two-part SSL prior to the
coefficients \(\boldsymbol{\beta}_j\). Specifically, we impose
independent group priors on the null space coefficients and on the
penalized space coefficients respectively,

\begin{align}\label{eq:bham_ssl}
  \beta^0_{j} |\gamma^0_{j},s_0,s_1 &\sim DE(0,(1-\gamma^0_{j}) s_0 + \gamma^0_{j} s_1) \nonumber \\
  \beta^\text{pen}_{jk} | \gamma^\text{pen}_{j},s_0,s_1 &\sim DE(0,(1-\gamma^\text{pen}_{j}) s_0 + \gamma^\text{pen}_{j} s_1), 
\end{align}

where \(\gamma^0_{j}\in\{0,1\}\) and
\(\gamma^\text{pen}_{j}\in \{0,1\}^{K_j}\) are latent indicator variable
and vectors, indicating if the model includes the linear effect and the
nonlinear effect of the \(j\)th variable respectively. \(s_0\) and
\(s_1\) are scale parameters, assuming \(0 < s_0 < s_1\) and given.
These scale parameters \(s_0\) and \(s_1\) can be treated as tuning
parameters and optimized via cross-validation. A discussion of how to
choose the scale parameters comes in Section \ref{sec:tune}. To note,
this prior differs from previous group spike-and-slab lasso priors
\citep{Tang2018, Tang2019}, where the \(\boldsymbol{\beta}^0_j\) and
\(\boldsymbol{\beta}^\text{pen}_j\) shares the same indicator variables
\(\gamma_j^0\), \(\gamma_j^\text{pen}\) respectively. It is possible to
add a more restrictive assumption on the priors, assuming that one
indicator variable decides the inclusion of both the linear effect and
nonlinear effect, i.e.~\(\gamma_j = \gamma^0_j = \gamma^\text{pen}_j\).
This converges to the SB-GAM \citep{Bai2021}. Conversely, it is also
possible to relax the assumption such that each coefficient \(\beta\) to
have its own latent indicator \(\gamma\), but at the cost of
complicating the bi-level function selection. This reduces the proposed
prior to the classic spike-and-slab lasso prior.

The re-parameterization introduced in Section \ref{sec:BHAM} grants the
validity of the proposed prior. First of all, the smoothing function
bases are linear dependent and ask for extra attention. The
eigeondecomposition remedies the problem and hence our prior can be set
to be conditionally independent. Secondly, the eigenvalue scaling
provides a panacea to allow unified scale parameters for all bases of
all smoothing functions of variables.
